---
title: "RHAU RNAseq Analysis: From Count Table to DEGs"
subtitle: "Paired RHAU vs WT workflow using Alpha-WT-KO-counts.csv"
author: "Joshua Howard"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This document analyzes the combined count table in
Alpha-WT-KO-counts.csv. The sample names suggest matched RHAU/WT pairs
by subject ID (12, 13, 14), so the DEG step uses a paired limma-voom
design with subject blocking.

# Required Packages

Install these once if needed.

```{r install-packages, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c("edgeR", "limma", "AnnotationDbi", "org.Mm.eg.db"))
install.packages(c("data.table", "ggplot2", "ggrepel", "ggfortify", "statmod"))
```

# Load Packages

```{r load-packages}
library(data.table)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(edgeR)
library(limma)
library(ggplot2)
library(ggrepel)
library(ggfortify)
```

# Read Count Table

```{r read-counts}
count_dir <- "//wsl.localhost/Ubuntu/home/unixuser/RHAU"
count_name <- "Alpha-WT-KO-counts.csv"

count_file <- file.path(count_dir, count_name)

count_raw <- fread(count_file, data.table = FALSE)

count_tbl <- count_raw

rowname_col <- "gene_id"
gene_ids <- count_tbl[[rowname_col]]
count_tbl[[rowname_col]] <- NULL
gene_symbols <- mapIds(
  org.Mm.eg.db,
  keys = gene_ids,
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

# FIX: If symbol is NA, keep the original Ensembl ID
gene_symbols <- ifelse(is.na(gene_symbols), gene_ids, gene_symbols)

count_tbl$gene_symbol <- gene_symbols

count_tbl <- aggregate(. ~ gene_symbol, data = count_tbl, FUN = sum, na.rm = TRUE, na.action = na.pass)

rownames(count_tbl) <- count_tbl$gene_symbol
count_tbl$gene_symbol <- NULL
```


# Create Metadata

```{r metadata}
sample_ids <- colnames(count_tbl)

meta <- data.frame(
  SampleID = sample_ids,
  Subject = sub("-(RHAU|WT)-Alpha$", "", sample_ids),
  Treatment = factor(
    ifelse(grepl("RHAU", sample_ids), "RHAU", "WT"),
    levels = c("WT", "RHAU")
  )
)

rownames(meta) <- meta$SampleID
meta$Subject <- factor(meta$Subject)
```

The sample names determine both treatment and subject pairing, so the
design stays aligned even if the columns are reordered.

# Build DGE Object

```{r dge-object}
dge <- DGEList(counts = count_tbl, samples = meta)
dge <- calcNormFactors(dge, method = "TMM")
```

# Filter Low-Expression Genes

```{r filter-low-expression}
keep <- filterByExpr(dge, group = meta$Treatment)
dge <- dge[keep, , keep.lib.sizes = FALSE]
```

This keeps genes with enough counts to support a reliable comparison.

# Normalization and PCA

```{r normalize-and-pca, fig.width=7, fig.height=6}
design <- model.matrix(~ 0 + Treatment, data = meta)
colnames(design) <- levels(meta$Treatment)

v0 <- voom(dge, design, plot = TRUE)
corfit <- duplicateCorrelation(v0, design, block = meta$Subject)
dge_v <- voom(dge, design, plot = TRUE, block = meta$Subject, correlation = corfit$consensus)

saveRDS(dge_v, "RHAU_dge_v.rds")

meta_table <- dge_v$targets
pca_prep <- prcomp(t(dge_v$E), scale. = TRUE)

pca_plot <- autoplot(
  pca_prep,
  data = meta_table,
  colour = "Treatment",
  shape = "Treatment",
) +
  geom_text_repel(aes(label = rownames(meta_table)), size = 4) +
  theme_classic()

pca_plot

ggsave("RHAU_PCA.pdf", plot = pca_plot, device = "pdf", units = "cm", width = 16, height = 14)
```

# Differential Expression Analysis

```{r deg-analysis}
comparison <- "RHAU-WT"
contrast_matrix <- makeContrasts(contrasts = comparison, levels = design)

fit <- lmFit(dge_v, design, block = meta$Subject, correlation = corfit$consensus)
fit_2 <- contrasts.fit(fit, contrast_matrix)
fit_2 <- eBayes(fit_2)

deg_tbl <- topTable(fit_2, coef = 1, n = Inf, p.value = 1, lfc = 0, sort.by = "P")
deg_tbl$gene_symbol <- rownames(deg_tbl)

fwrite(deg_tbl, "RHAU_WT_DEG_results.tsv", sep = "\t", row.names = TRUE)

head(deg_tbl)
```

Positive logFC values mean higher expression in RHAU relative to WT.
If you want the reverse biological direction, change the contrast to
WT-RHAU.

# Volcano Plot

```{r volcano-plot, fig.width=8, fig.height=7}
deg_tbl$neg_log10_adj_p <- -log10(pmax(deg_tbl$adj.P.Val, .Machine$double.xmin))
deg_tbl$direction <- with(
  deg_tbl,
  ifelse(adj.P.Val < 0.05 & logFC >= 1, "Upregulated",
         ifelse(adj.P.Val < 0.05 & logFC <= -1, "Downregulated", "Not significant"))
)

top_up_df <- deg_tbl[deg_tbl$direction == "Upregulated", ][order(deg_tbl[deg_tbl$direction == "Upregulated", "adj.P.Val"]), ]
top_down_df <- deg_tbl[deg_tbl$direction == "Downregulated", ][order(deg_tbl[deg_tbl$direction == "Downregulated", "adj.P.Val"]), ]
label_df <- rbind(
  head(top_up_df, 15),
  head(top_down_df, 15)
)

volcano_plot <- ggplot(deg_tbl, aes(x = logFC, y = neg_log10_adj_p)) +
  geom_point(aes(color = direction), alpha = 0.8, size = 1.8) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  geom_text_repel(data = label_df, aes(label = gene_symbol), size = 3, max.overlaps = 20) +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Not significant" = "grey70")) +
  labs(
    title = "RHAU vs WT volcano plot",
    x = "log2 fold change",
    y = "-log10 adjusted p-value",
    color = "Direction"
  ) +
  theme_classic()

volcano_plot

ggsave("RHAU_WT_Volcano.pdf", plot = volcano_plot, device = "pdf", units = "cm", width = 18, height = 15)
```

# Top Differentially Expressed Genes

```{r top-deg-table, results='asis'}
top_up_tbl <- head(top_up_df[, c("gene_symbol", "logFC", "adj.P.Val")], 15)
top_down_tbl <- head(top_down_df[, c("gene_symbol", "logFC", "adj.P.Val")], 15)

top_deg_tbl <- rbind(
  transform(top_up_tbl, Direction = "Upregulated"),
  transform(top_down_tbl, Direction = "Downregulated")
)

top_deg_tbl <- top_deg_tbl[, c("Direction", "gene_symbol", "logFC", "adj.P.Val")]
colnames(top_deg_tbl) <- c("Direction", "Gene Symbol", "logFC", "adj.P.Val")

knitr::kable(top_deg_tbl, digits = 4, caption = "Top 15 upregulated and top 15 downregulated genes")
```

# Interpretation

The files saved by this workflow are:

- RHAU_dge_v.rds for the normalized voom object
- RHAU_PCA.pdf for the sample clustering plot
- RHAU_WT_DEG_results.tsv for the full DEG table
- RHAU_WT_Volcano.pdf for the volcano plot

If your experiment is not truly paired, remove the Subject block from the
voom and lmFit steps and use a standard two-group design instead.
