---
title: "RHAU RNAseq Analysis: From Count Table to DEGs"
subtitle: "Paired RHAU vs WT workflow using Alpha-WT-KO-counts.csv"
author: "Joshua Howard"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This document analyzes the combined count table in
Alpha-WT-KO-counts.csv. The sample names suggest matched RHAU/WT pairs
by subject ID (12, 13, 14), so the DEG step uses a paired limma-voom
design with subject blocking. Pathway analysis (GSEA and ORA against
MSigDB gene sets) is run at the end on the resulting DEGs.

# Required Packages

Install these once if needed.

```{r install-packages, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c("edgeR", "limma", "AnnotationDbi", "org.Mm.eg.db",
                        "clusterProfiler", "enrichplot", "rtracklayer"))
install.packages(c("data.table", "ggplot2", "ggrepel", "ggfortify", "statmod", "pheatmap", "dplyr"))
```

# Load Packages

```{r load-packages}
library(data.table)
library(AnnotationDbi)
library(org.Mm.eg.db)
library(edgeR)
library(limma)
library(ggplot2)
library(ggrepel)
library(ggfortify)
library(pheatmap)
library(dplyr)
library(clusterProfiler)
library(enrichplot)
library(rtracklayer)
```

# Read Count Table

```{r read-counts}
count_dir <- "//wsl.localhost/Ubuntu/home/unixuser/RHAU"
count_name <- "Alpha-WT-KO-counts.csv"
count_file <- file.path(count_dir, count_name)

# 1. Load your count data
count_raw <- fread(count_file, data.table = FALSE)
count_tbl <- count_raw

rowname_col <- "gene_id"
gene_ids <- count_tbl[[rowname_col]]
count_tbl[[rowname_col]] <- NULL

# 2. Load and parse the GTF file
# (Adjust the path and filename to match your actual GTF file)
gtf_name <- "gencode.vM33.annotation.gtf" 
gtf_file <- file.path(count_dir, gtf_name)

# Import GTF and convert to a data frame
gtf_data <- as.data.frame(import(gtf_file))

# Filter for gene features to speed up mapping
gene_map <- subset(gtf_data, type == "gene")

# 3. Map Ensembl IDs to Symbols using the GTF data
# Creates a named vector: names = gene_id, values = gene_name
# Note: GTF files typically use 'gene_id' and 'gene_name' (symbol)
ensembl_to_symbol <- setNames(gene_map$gene_name, gene_map$gene_id)

# Match your count IDs to the GTF mapping vector
gene_symbols <- ensembl_to_symbol[gene_ids]

# 4. If symbol is NA, keep the original Ensembl ID
gene_symbols <- ifelse(is.na(gene_symbols), gene_ids, gene_symbols)

# 5. Attach symbols and aggregate duplicates
count_tbl$gene_symbol <- gene_symbols
count_tbl <- aggregate(. ~ gene_symbol, data = count_tbl, FUN = sum, na.rm = TRUE, na.action = na.pass)

rownames(count_tbl) <- count_tbl$gene_symbol
count_tbl$gene_symbol <- NULL
```



# Create Metadata

```{r metadata}
sample_ids <- colnames(count_tbl)

meta <- data.frame(
  SampleID = sample_ids,
  Subject = sub("-(RHAU|WT)-Alpha$", "", sample_ids),
  Treatment = factor(
    ifelse(grepl("RHAU", sample_ids), "RHAU", "WT"),
    levels = c("WT", "RHAU")
  )
)

rownames(meta) <- meta$SampleID
meta$Subject <- factor(meta$Subject)

# Consistent color mapping used across all plots below.
# Named by factor level so ggplot always matches the right color to the
# right group, regardless of row order or how a data frame gets sorted.
treatment_colors <- c(WT = "blue", RHAU = "red")
```

The sample names determine both treatment and subject pairing, so the
design stays aligned even if the columns are reordered.

# Build DGE Object

```{r dge-object}
dge <- DGEList(counts = count_tbl, samples = meta)
dge <- normLibSizes(dge, method = "TMM")
```

# Filter Low-Expression Genes

```{r filter-low-expression}
keep <- filterByExpr(dge, group = meta$Treatment, min.count = 1, min.total.count = 3)
dge <- dge[keep, , keep.lib.sizes = FALSE]
```

This keeps genes with enough counts to support a reliable comparison.

# Normalization and PCA

```{r normalize-and-pca, fig.width=7, fig.height=6}
design <- model.matrix(~ 0 + Treatment, data = meta)
colnames(design) <- levels(meta$Treatment)

v0 <- voom(dge, design, plot = TRUE)
corfit <- duplicateCorrelation(v0, design, block = meta$Subject)
dge_v <- voom(dge, design, plot = TRUE, block = meta$Subject, correlation = corfit$consensus)

saveRDS(dge_v, "RHAU_dge_v.rds")

meta_table <- dge_v$targets
pca_prep <- prcomp(t(dge_v$E), scale. = TRUE)

pca_plot <- autoplot(
  pca_prep,
  data = meta_table,
  colour = "Treatment",
  shape = "Treatment",
) +
  scale_color_manual(values = c("blue", "red")) +
  geom_text_repel(aes(label = rownames(meta_table)), size = 4) +
  theme_classic()

pca_plot

ggsave("RHAU_PCA.pdf", plot = pca_plot, device = "pdf", units = "cm", width = 16, height = 14)
```

# Differential Expression Analysis

```{r deg-analysis}
comparison <- "RHAU-WT"
contrast_matrix <- makeContrasts(contrasts = comparison, levels = design)

fit <- lmFit(dge_v, design, block = meta$Subject, correlation = corfit$consensus)
fit_2 <- contrasts.fit(fit, contrast_matrix)
fit_2 <- eBayes(fit_2)

deg_tbl <- topTable(fit_2, coef = 1, n = Inf, p.value = 1, lfc = 0, sort.by = "P")
deg_tbl$gene_symbol <- rownames(deg_tbl)

fwrite(deg_tbl, "RHAU_WT_DEG_results.tsv", sep = "\t", row.names = TRUE)

head(deg_tbl)
```

Positive logFC values mean higher expression in RHAU relative to WT.
If you want the reverse biological direction, change the contrast to
WT-RHAU.

# Volcano Plot

```{r volcano-plot, fig.width=8, fig.height=7}
deg_tbl$neg_log10_adj_p <- -log10(pmax(deg_tbl$adj.P.Val, .Machine$double.xmin))
deg_tbl$direction <- with(
  deg_tbl,
  ifelse(adj.P.Val < 0.05 & logFC >= 1, "Upregulated",
         ifelse(adj.P.Val < 0.05 & logFC <= -1, "Downregulated", "Not significant"))
)

top_up_df <- deg_tbl[deg_tbl$direction == "Upregulated", ][order(deg_tbl[deg_tbl$direction == "Upregulated", "adj.P.Val"]), ]
top_down_df <- deg_tbl[deg_tbl$direction == "Downregulated", ][order(deg_tbl[deg_tbl$direction == "Downregulated", "adj.P.Val"]), ]
label_df <- rbind(
  head(top_up_df, 15),
  head(top_down_df, 15)
)

volcano_plot <- ggplot(deg_tbl, aes(x = logFC, y = neg_log10_adj_p)) +
  geom_point(aes(color = direction), alpha = 0.8, size = 1.8) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  geom_text_repel(data = label_df, aes(label = gene_symbol), size = 3, max.overlaps = 20) +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Not significant" = "grey70")) +
  labs(
    title = "RHAU vs WT volcano plot",
    x = "log2 fold change",
    y = "-log10 adjusted p-value",
    color = "Direction"
  ) +
  theme_classic()

volcano_plot

ggsave("RHAU_WT_Volcano.pdf", plot = volcano_plot, device = "pdf", units = "cm", width = 18, height = 15)
```

# Top Differentially Expressed Genes

```{r top-deg-table, results='asis'}
top_up_tbl <- head(top_up_df[, c("logFC", "adj.P.Val", "AveExpr")], 15)
top_down_tbl <- head(top_down_df[, c("logFC", "adj.P.Val", "AveExpr")], 15)

top_deg_tbl <- rbind(
  transform(top_up_tbl, Direction = "Upregulated"),
  transform(top_down_tbl, Direction = "Downregulated")
)

top_deg_tbl <- top_deg_tbl[, c("Direction", "logFC", "adj.P.Val", "AveExpr")]
colnames(top_deg_tbl) <- c("Direction", "logFC", "adj.P.Val", "AveExpr")

knitr::kable(top_deg_tbl, digits = 4, caption = "Top 15 upregulated and top 15 downregulated genes")
```

# DEG Heatmap

```{r heatmap, fig.width=6, fig.height=6}
# Significance thresholds for which DEGs go into the heatmap.
# These are independent of the fixed |logFC| >= 1 / adj.P < 0.05 lines
# drawn on the volcano plot above, so tune them here as needed.
FC_threshold <- 2      # linear fold-change cutoff (log2(FC_threshold) is applied below)
P_threshold  <- 0.05   # adjusted p-value cutoff

# Prepare sample annotations, ordered so columns group by treatment
col_annot_df <- data.frame(
  Treatment = dge_v$targets$Treatment,
  row.names = rownames(dge_v$targets)
)
col_annot_df <- col_annot_df[order(col_annot_df$Treatment), , drop = FALSE]

# Select significant DEGs
# NOTE: deg_tbl needs a comma to subset rows (not columns), and the gene
# names live in the gene_symbol column, not V1.
deg_sig <- deg_tbl[
  abs(deg_tbl$logFC) > log2(FC_threshold) & deg_tbl$adj.P.Val < P_threshold,
  "gene_symbol"
]

if (length(deg_sig) < 2) {
  warning("Fewer than 2 genes pass the heatmap thresholds (FC_threshold = ",
          FC_threshold, ", P_threshold = ", P_threshold,
          "). Consider relaxing these before plotting.")
}

expr_deg <- dge_v$E[deg_sig, rownames(col_annot_df), drop = FALSE]

# Create heatmap
heatmap_custom <- pheatmap(
  expr_deg,
  scale = "row",
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  show_colnames = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  annotation_col = col_annot_df,
  annotation_colors = list(Treatment = treatment_colors),
  main = "Differential Expression Heatmap"
)

ggsave("RHAU_WT_Heatmap.pdf", plot = heatmap_custom, device = "pdf", units = "cm", width = 16, height = 18)
```

Rows are z-scored (`scale = "row"`) so the color scale reflects each
gene's relative expression across samples, not its absolute expression
level. Columns are grouped by treatment rather than clustered, so the
WT/RHAU split stays easy to read.

# Gene Panel Expression Heatmap

```{r heatmap-legend, fig.width=8, fig.height=8}
# DEG Heatmap — CSR + DNA Repair + Proliferation gene panel
# Rows ordered and grouped by gene function (CSR vs DNA Repair vs Proliferation), not clustered,
# so the three groups form clean, contiguous blocks.

csr_genes <- c(
  # DNA Cleavage & Initiation
  "Aicda", "Ung", "Smug1", "Apex1"
)

dna_repair_genes <- c(
  # NHEJ Repair
  "Xrcc6", "Xrcc5", "Prkdc", "Lig4", "Xrcc4", "Nhej1", "Paxx",
  # A-EJ Repair
  "Parp1", "Lig3", "Xrcc1",
  # DNA Damage Response & Synapsis
  "Atm", "Trp53bp1", "Mre11", "Rad50", "Nbn", "Stag2", "Smc1a", "Smc3"
)

proliferation_genes <- c(
  # Cell Cycle Entry & Checkpoints
  "Myc", "Ccnd2", "Ccnd3", "Cdk4", "Cdk6", "E2f1",
  # DNA Replication & Elongation (S-Phase)
  "Pcna", "Mki67", "Mcm2", "Mcm3", "Mcm4", "Mcm5", "Mcm6", "Mcm7",
  # Upstream Proliferative Signaling Receptors
  "Cd40", "Tnfrsf13b", "Tnfrsf17", "Il4r"
)

transcription_factors <- c(
  # B Cell Lineage & Differentiation
  "Tbx21", "Bcl6", "Prdm1", "Irf4", "Irf8", "Spib", "Spi1", "Ebf1"
)

gene_panel <- data.frame(
  gene_symbol = c(csr_genes, dna_repair_genes, proliferation_genes, transcription_factors),
  GeneGroup = c(rep("CSR", length(csr_genes)),
                rep("DNA Repair", length(dna_repair_genes)),
                rep("Proliferation", length(proliferation_genes)),
                rep("Transcription Factor", length(transcription_factors))),
  stringsAsFactors = FALSE
)

# Match against the filtered/voom-normalized gene set
available_genes <- rownames(dge_v$E)
found <- gene_panel$gene_symbol %in% available_genes
missing_genes <- gene_panel$gene_symbol[!found]

if (length(missing_genes) > 0) {
  warning("The following genes were not found in dge_v$E (filtered out earlier, ",
          "or symbol mismatch): ", paste(missing_genes, collapse = ", "))
}

gene_panel <- gene_panel[found, ]

if (nrow(gene_panel) < 2) {
  warning("Fewer than 2 panel genes were found in the filtered dataset. ",
          "Check symbol spelling and the low-expression filter above.")
}

# Keep CSR genes first, DNA Repair genes second, Proliferation genes third, preserving the functional
# sub-order within each group (rather than alphabetical or cluster-driven order)
gene_panel$GeneGroup <- factor(gene_panel$GeneGroup, levels = c("CSR", "DNA Repair"
                                                                , "Proliferation", "Transcription Factor"
                                                                ))
gene_panel <- gene_panel[order(gene_panel$GeneGroup), ]

# Row index where the CSR block ends, used to draw a gap line between blocks
csr_block_size <- sum(gene_panel$GeneGroup == "CSR")

# Prepare sample annotations, ordered so columns group by treatment
col_annot_df <- data.frame(
  Treatment = dge_v$targets$Treatment,
  row.names = rownames(dge_v$targets)
)
col_annot_df <- col_annot_df[order(col_annot_df$Treatment), , drop = FALSE]

# Row annotation by gene functional group
row_annot_df <- data.frame(
  "Gene Group" = gene_panel$GeneGroup,
  row.names = gene_panel$gene_symbol,
  check.names = FALSE
)

group_colors <- c(CSR = "black", "DNA Repair" = "red"
                  , Proliferation = "white", "Transcription Factor" = "blue"
                  )

expr_deg <- dge_v$E[gene_panel$gene_symbol, rownames(col_annot_df), drop = FALSE]

# Create heatmap
heatmap_custom <- pheatmap(
  expr_deg,
  scale = "row",
  treeheight_row = 0, 
  treeheight_col = 0,
  cluster_rows = TRUE,   # preserve CSR / Proliferation block order instead of clustering
  cluster_cols = FALSE,
  gaps_row = csr_block_size,  # draws a visual line between the two gene groups
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 7,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  annotation_col = col_annot_df,
  annotation_row = row_annot_df,
  annotation_colors = list(Treatment = treatment_colors, "Gene Group" = group_colors),
  main = "Gene Group Expression Heatmap"
)

ggsave("RHAU_CSR_Proliferation_Heatmap.pdf", plot = heatmap_custom, device = "pdf",
       units = "cm", width = 24, height = 24)
```

# Individual Gene Expression

```{r gene-setup}
gene_to_plot <- "Tbx21"  # Replace with your gene of interest

if (!gene_to_plot %in% rownames(dge_v$E)) {
  stop(gene_to_plot, " was not found in dge_v$E. Check spelling/capitalization, ",
       "or confirm it survived the low-expression filtering step above.")
}

# Transform all gene expression values to positive (needed for some plot types)
dge_v2 <- dge_v
dge_v2$E <- dge_v$E + abs(min(dge_v$E)) + 1

# dge_v$E holds voom log2-CPM values, which are legitimately negative for
# lowly expressed genes -- that's expected and not shifted here, since
# doing so would distort the log2 fold-change scale for every downstream
# use of dge_v$E (including the heatmap above).
gene_data <- data.frame(
  Treatment = dge_v$targets$Treatment,
  Expression = dge_v$E[gene_to_plot, ]
)

# Keep treatment order stable in the plots
gene_data$Treatment <- factor(gene_data$Treatment, levels = levels(meta$Treatment))

# Quick sanity check: inspect the mean expression by group before interpreting direction
gene_summary <- gene_data %>%
  group_by(Treatment) %>%
  summarise(mean_expression = mean(Expression), .groups = "drop")

gene_summary
```


```{r box-plot, fig.width=5, fig.height=5}
boxplot <- ggplot(gene_data, aes(x = Treatment, y = Expression)) +
  geom_boxplot(aes(fill = Treatment), alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  scale_fill_manual(values = treatment_colors) +
  theme_minimal() +
  labs(title = paste(gene_to_plot, "Expression"), y = "log2 CPM (voom)") +
  theme(legend.position = "none")

boxplot

ggsave(paste0("RHAU_", gene_to_plot, "_Boxplot.pdf"), plot = boxplot, device = "pdf", units = "cm", width = 12, height = 12)
```

# MSigDB Pathway Analysis

The most comprehensive option — incorporates GO, KEGG, Hallmark,
Reactome, and other pathway collections in one analysis.

Download the appropriate gene set `.gmt` file from the MSigDB website
(choose the species-specific file — mouse or human). The chunk below
assumes it has been placed in the **same directory as the count file**
(`count_dir`), consistent with the rest of this document. Update
`gmt_name` if your filename differs (MSigDB version numbers change
periodically).

```{r gmt-setup}
gmt_name <- "msigdb.v2025.1.Mm.symbols.gmt"
gmt_file <- file.path(count_dir, gmt_name)

if (!file.exists(gmt_file)) {
  stop("GMT file not found at ", gmt_file,
       ". Confirm the filename/version and that it was downloaded into count_dir.")
}

msigdb <- read.gmt(gmt_file)
```

## Rank Genes for GSEA

GSEA needs a numeric vector of every tested gene, named by gene symbol
and sorted from highest to lowest. Ranking by logFC alone is used here
for consistency with the volcano/heatmap logic above; ranking by a
signed significance statistic (e.g. `-log10(P.Value) * sign(logFC)`)
is a common alternative if you want the ranking to also weight
significance, not just effect size.

```{r rank-genes}
logfc <- deg_tbl$logFC
names(logfc) <- deg_tbl$gene_symbol

# Drop any non-finite values and enforce the decreasing order GSEA expects
logfc <- logfc[is.finite(logfc)]
logfc <- sort(logfc, decreasing = TRUE)
```

## Define Up/Down Gene Sets for ORA

ORA (over-representation) needs a simple gene list rather than a
ranked vector, and ideally a defined background/`universe` — here, all
genes that survived low-expression filtering — so enrichment is judged
against genes that could plausibly have been detected as a DEG, not
against the whole genome.

```{r ora-gene-sets}
deg_up   <- deg_tbl$gene_symbol[deg_tbl$direction == "Upregulated"]
deg_down <- deg_tbl$gene_symbol[deg_tbl$direction == "Downregulated"]

universe <- rownames(dge_v$E)
```

## GSEA Method

```{r msigdb-gsea}
enrich_msigdb_gsea <- GSEA(logfc, TERM2GENE = msigdb, pvalueCutoff = 0.05)

enrich_msigdb_gsea_df <- enrich_msigdb_gsea@result

fwrite(enrich_msigdb_gsea_df, "enrich_msigdb_gsea_df.tsv", sep = "\t")

head(enrich_msigdb_gsea_df)
```

```{r msigdb-gsea-plot, fig.width=8, fig.height=6}
if (nrow(enrich_msigdb_gsea_df) > 0) {
  dotplot(enrich_msigdb_gsea, showCategory = 8, font.size = 8, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA: MSigDB pathways enriched by rank (logFC)")
}
```

## ORA Method

Run separately for up- and down-regulated genes so the direction of
each enriched pathway is unambiguous.

```{r msigdb-ora}
enrich_msigdb_fet_up <- enricher(deg_up, TERM2GENE = msigdb, universe = universe, pvalueCutoff = 1)
enrich_msigdb_fet_down <- enricher(deg_down, TERM2GENE = msigdb, universe = universe, pvalueCutoff = 1)

enrich_msigdb_fet_up_df <- enrich_msigdb_fet_up@result
enrich_msigdb_fet_down_df <- enrich_msigdb_fet_down@result

fwrite(enrich_msigdb_fet_up_df, "enrich_msigdb_fet_up_df.tsv", sep = "\t")
fwrite(enrich_msigdb_fet_down_df, "enrich_msigdb_fet_down_df.tsv", sep = "\t")

head(enrich_msigdb_fet_up_df)
head(enrich_msigdb_fet_down_df)
```

```{r msigdb-ora-plot, fig.width=8, fig.height=6}
if (nrow(enrich_msigdb_fet_up_df) > 0) {
  dotplot(enrich_msigdb_fet_up, showCategory = 15) + ggtitle("ORA: pathways enriched in upregulated genes")
}
if (nrow(enrich_msigdb_fet_down_df) > 0) {
  dotplot(enrich_msigdb_fet_down, showCategory = 15) + ggtitle("ORA: pathways enriched in downregulated genes")
}
```

GSEA uses the full ranked gene list and detects coordinated shifts
even when individual genes don't cross a significance threshold; ORA
only considers genes that already passed the DEG cutoff, so it's more
conservative but easier to interpret directionally (separate up/down
results). Reporting both is common practice — treat agreement between
the two as the strongest evidence for a pathway.

# Interpretation

The files saved by this workflow are:

- RHAU_dge_v.rds for the normalized voom object
- RHAU_PCA.pdf for the sample clustering plot
- RHAU_WT_DEG_results.tsv for the full DEG table
- RHAU_WT_Volcano.pdf for the volcano plot
- RHAU_WT_Heatmap.pdf for the DEG heatmap
- RHAU_\<gene\>_Boxplot.pdf for the individual gene expression plot
- enrich_msigdb_gsea_df.tsv for the full MSigDB GSEA results
- enrich_msigdb_fet_up_df.tsv / enrich_msigdb_fet_down_df.tsv for the
  MSigDB ORA results, split by direction
