---
title: "RNAseq Analysis: From Count Table to DEGs"
subtitle: "Beginner's Workflow (Part 3) — Best Practices"
author: "Dr. Lei Guo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This workflow picks up after you have generated a gene expression count
matrix from raw sequencing data (see Parts 1 and 2 of the series, which
cover processing FASTQ files into a count matrix). Here we move from the
Linux environment into R to perform differential expression (DE)
analysis, data visualization, and other downstream statistical analyses.

# Installing R and RStudio

Installation is done outside of this script:

1.  **Install R** — download the latest version for your OS from the
    official R website and follow the installer prompts.
2.  **Install RStudio** — download the appropriate version from the
    RStudio download page and install it like any other application.

RStudio provides:

- **Code Editor Window** — write/edit scripts; run selected code with
  "Run" or the whole script with "Source".
- **R Console Window** — real-time interaction with R; shows output,
  errors, and print statements.
- **Work Space Window** — lists all objects in your current session
  (variables, data sets, functions).
- **Plot & Files Window** — view, export, or zoom into generated plots.

# Setting Up Your R Environment

## Installing Required Packages

Packages come from two main repositories: CRAN and Bioconductor. Run
this once to install everything needed for the workflow below.

```{r install-packages, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c("sva", "edgeR", "limma", "Biobase", "biomaRt",
                        "clusterProfiler", "EnhancedVolcano", "org.Hs.eg.db",
                        "org.Mm.eg.db", "org.Rn.eg.db"))

install.packages(c("data.table", "readxl", "stringr", "ggplot2", "ggrepel",
                    "ggfortify", "ggprism", "pheatmap", "VennDiagram",
                    "corrplot", "Hmisc", "stats", "tidyverse"))
```

If prompted to update existing packages, you can press "N" to skip.

## Loading Installed Packages

```{r load-packages}
library(data.table)
library(edgeR)
library(limma)
library(ggplot2)
library(ggrepel)
library(ggfortify)
library(stats)
library(sva)
```

## Setting Your Working Directory

Create a project folder on your computer and set it as the working
directory. All outputs will be saved here unless otherwise specified.

```{r set-wd, eval=FALSE}
# MacOS
# setwd("/Users/yourusername/Documents/RNAseq")

# Windows
setwd("//wsl.localhost/Ubuntu/home/unixuser/rnaseq_output/counts/")
```

Replace `yourusername` with your actual username.

# Data Preparation and Analysis

## Reading Data

### Building the count matrix from featureCounts output (optional)

If you haven't yet combined individual count files into a single
expression matrix, do so as follows:

```{r build-count-matrix, eval=FALSE}
# Retrieve count file paths
count_file_path <- ("//wsl.localhost/Ubuntu/home/unixuser/rnaseq_output/counts/")
file_names <- list.files(count_file_path,
                          pattern = "featureCounts_exon\\.txt$",
                          recursive = TRUE, full.names = TRUE)

# Read all count files into a DGEList
dgls <- readDGE(file_names, columns = c(1, 7), skip = 1)
counts_raw <- as.data.frame(dgls$counts)

# Rename samples
sample_name <- gsub("_featureCounts_exon", "", basename(colnames(counts_raw)))
colnames(counts_raw) <- sample_name

# Save the combined count table
fwrite(counts_raw, "counts_raw.tsv", sep = "\t")
```

### Reading the combined count table

```{r read-counts, eval=FALSE}
count_tbl <- fread("counts_raw.tsv", data.table = FALSE)
```

View it via `View(count_tbl)` in the console, or by clicking `count_tbl`
in the Work Space window.

## Data Cleaning

### Add row names to the count table

```{r rownames}
rownames(count_tbl) <- rownames(counts_raw)
#Old code do not use
#rownames(count_tbl) <- count_tbl[[1]]
#count_tbl <- count_tbl[, -1]
```

> **Note:** If your table has separate `GeneID` and `GeneSymbol` columns
> (rather than a single identifier column), see the [Handling a GeneID +
> GeneSymbol annotation
> table](#handling-a-geneid-genesymbol-annotation-table) section below
> before proceeding, to avoid duplicate row name errors.

### Filtering low-expressed genes

```{r filter-low-expr}
perc_keep <- 0.8
gene_keep <- rowSums(count_tbl > 0) >= ceiling(perc_keep * ncol(count_tbl))
count_tbl_low_rm <- count_tbl[gene_keep, ]
```

This keeps genes that are expressed (count \> 0) in at least 80% of
samples.

## Creating Metadata

```{r metadata}
meta <- data.frame(SampleID = colnames(count_tbl),
                    Treatment = c("KRAS_SPIB", "KRAS_SPIB", "KRAS", "KRAS"))
rownames(meta) <- meta$SampleID
```

Update `Treatment` to match your actual sample groups and ensure the
length matches `ncol(count_tbl)`.

## Creating a DGE List

```{r dge-list}
dge <- DGEList(counts = count_tbl_low_rm, samples = meta)
```

## Normalization

```{r normalize, fig.cap="Mean-variance trend after voom normalization"}
dge <- calcNormFactors(dge, method = "TMM")
dge_v <- voom(dge, plot = TRUE)

# Save the dge_v object for later use
saveRDS(dge_v, "dge_v.rds")
```

## Creating a PCA Plot

A PCA plot visualizes the overall structure of the data and helps
identify sample clustering, batch effects, or outliers. Samples from the
same treatment group should tend to cluster together.

```{r pca-plot, fig.width=7, fig.height=6}
shape_column <- "Treatment"
color_column <- "Treatment"
label <- TRUE
label_size <- 4
plot_save_name <- "PCA_Plot.pdf"

meta_table <- dge_v$targets
count_table_t <- as.data.frame(t(dge_v$E))
pca_prep <- prcomp(count_table_t, scale. = TRUE)

pca_plot <- autoplot(pca_prep, label, shape = shape_column,
                      data = meta_table, colour = color_column) +
  geom_text_repel(aes(label = rownames(meta_table)), size = label_size) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank())

pca_plot

ggsave(plot_save_name, plot = pca_plot, device = "pdf",
       units = "cm", width = 16, height = 14)
```

# Differential Expression Analysis

Perform DE analysis using `limma`. **Special characters are not allowed
in sample group names.**

```{r deg-analysis}
comparison <- "KRAS_SPIB-KRAS"

design <- model.matrix(~ 0 + dge_v$targets[["Treatment"]])
colnames(design) <- gsub(".*]]", "", colnames(design))

contrast_matrix <- makeContrasts(contrasts = comparison, levels = design)

fit <- lmFit(dge_v, design)
fit_2 <- contrasts.fit(fit, contrast_matrix)
fit_2 <- eBayes(fit_2)

deg_tbl <- topTable(fit_2, coef = colnames(contrast_matrix), n = Inf,
                     p.value = 1, lfc = 0, sort.by = "p")

fwrite(deg_tbl, "DEG_P1_lfc0.tsv", sep = "\t", row.names = TRUE)

head(deg_tbl)
```

This compares the `KRAS_SPIB` group to the `KRAS` group. Results are
stored in `deg_tbl` and saved to `DEG_P1_lfc0.tsv` in the working
directory.

**Column definitions:**

| Column | Meaning |
|------------------------------------|------------------------------------|
| `logFC` | log2 Fold Change |
| `AveExpr` | Average expression level of the gene across all samples |
| `t` | t-statistic — how strongly the gene is differentially expressed between conditions |
| `P.Value` | p-value testing the null hypothesis of no difference in expression between groups |
| `adj.P.Val` | p-value adjusted for multiple testing |
| `B` | B-statistic (log-odds) — log-odds that a gene is differentially expressed |

# Handling a GeneID + GeneSymbol annotation table {#handling-a-geneid-genesymbol-annotation-table}

If your count table has both a `GeneID` and a `GeneSymbol` column as the
first two columns (and simply using `GeneSymbol` as row names causes a
`duplicate 'row.names' are not allowed` error), use this approach
instead:

```{r geneid-genesymbol, eval=FALSE}
# make a gene annotation table
annot_tbl <- data.frame(GeneID = count_tbl[[1]], GeneSymbol = count_tbl[[2]])
rownames(annot_tbl) <- annot_tbl[[1]]

# remove the "GeneID" and "GeneSymbol" columns from your count table
count_tbl <- count_tbl[, -c(1, 2)]

# filter out low-expressed genes the same way as above
gene_keep <- rowSums(count_tbl > 0) >= ceiling(perc_keep * ncol(count_tbl))
count_tbl_low_rm <- count_tbl[gene_keep, ]

# update the annotation table to match
annot_tbl <- annot_tbl[rownames(count_tbl_low_rm), ]

# create metadata table the same way as above

# create a DGEList with the annotation table included
dge <- DGEList(counts = count_tbl_low_rm, genes = annot_tbl, samples = meta)

# the rest of the steps are the same
```

# Troubleshooting Notes

- **`Error in .rowNamesDF<-(x, value = value): duplicate 'row.names' are not allowed`**
  — usually caused by duplicate gene symbols when an extra `GeneID`
  column is present. See [Handling a GeneID + GeneSymbol annotation
  table](#handling-a-geneid-genesymbol-annotation-table).
- **`Error in taglist[[i]] : subscript out of bounds`** during
  `readDGE()` — your featureCounts tables likely have a different column
  structure than expected (columns 1 and 7, after skipping the first
  row). Reformat your tables to match the expected structure.
- **`Setting row names on a tibble is deprecated` / mismatched row
  counts when building metadata** — make sure you're reading the count
  table with `fread(..., data.table = FALSE)` (a base data frame), not
  leaving it as a tibble, and that `SampleID` and `Treatment` in `meta`
  have matching lengths.
- **Low count filtering** — when filtering for a minimum count
  threshold, apply it row-wise across samples (e.g.
  `rowSums(counts >= 10) == ncol(counts)`), not by subsetting a single
  row index. A construct like `dds[keep, ]` where `keep` is a scalar
  (e.g. `keep = 10`) will incorrectly keep only that one row.

# Conclusion

You've completed the differential expression analysis of your RNAseq
data. The resulting list of differentially expressed genes (DEGs) opens
up many possibilities for deeper exploration, including functional
enrichment analysis. Future steps typically include interpreting these
results (e.g. volcano plots, GSEA/ORA pathway enrichment) and further
downstream biological analyses.

# References

Burger T. Fudging the volcano-plot without dredging the data. *Nat
Commun.* 2024 Feb 15;15(1):1392. doi: 10.1038/s41467-024-45834-7. PMID:
38360828; PMCID: PMC10869345.
