---
title: "RNAseq Analysis: Creating Publication-Ready Figures"
subtitle: "Beginner's Workflow (Part 4) — Volcano Plots, Heatmaps, and Gene-Level Visualizations"
author: "Dr. Lei Guo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

After identifying differentially expressed genes (DEGs) in Part 3, the
next step is visualizing the results effectively. Good visualizations
turn complex genomic data into clear, interpretable figures that reveal
patterns and help communicate findings to both technical and
non-technical audiences.

This workflow covers:

- Volcano plots (two approaches)
- Heatmaps of significant DEGs
- Individual gene expression plots (violin, box, and bar plots)

# Essential R Packages for Data Visualization

If you followed Part 3, these packages should already be installed. If
not, see the "Installing Required Packages" section of that tutorial.

```{r load-packages}
library(data.table)      # For efficient data handling
library(tidyverse)       # For versatile data manipulation
library(ggplot2)         # For creating custom plots
library(ggrepel)         # For non-overlapping text labels
library(ggfortify)       # For statistical visualizations
library(ggprism)         # For publication-ready styling
library(pheatmap)        # For creating heatmaps
library(EnhancedVolcano) # For volcano plots
```

If you want to use the ComplexHeatmap section below, install these
packages first:

```{r install-complexheatmap, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install(c("ComplexHeatmap", "circlize"))
```

# Loading Differential Expression Results

These files (`DEG_P1_lfc0.tsv` and `dge_v.rds`) were generated at the
end of Part 3.

```{r load-data}
# Read differential expression results
deg_tbl <- fread("DEG_P1_lfc0.tsv")
dge_v <- readRDS("dge_v.rds")

# Set significance thresholds
FC_threshold <- 2
P_threshold <- 0.05
```

# Creating Volcano Plots

## Method 1: Quick Visualization with EnhancedVolcano

The `EnhancedVolcano` package offers a fast way to build informative
volcano plots. It has some limitations with gene label placement, but
it's great for a quick overview of your DE results.

### Set up the color scheme

```{r volcano-colors}
# Create color scheme for different gene categories
keyvals.colour_rd <- ifelse(
  (deg_tbl$logFC < -log2(FC_threshold) & deg_tbl$adj.P.Val < P_threshold), 'blue',
  ifelse((deg_tbl$logFC > log2(FC_threshold) & deg_tbl$adj.P.Val < P_threshold), 'red', 'grey30')
)

# Add informative labels showing number of genes in each category
names(keyvals.colour_rd)[keyvals.colour_rd == 'blue'] <- paste0(
  "Downregulated (n=",
  sum(deg_tbl$logFC < -log2(FC_threshold) & deg_tbl$adj.P.Val < P_threshold),
  ")"
)

names(keyvals.colour_rd)[keyvals.colour_rd == 'red'] <- paste0(
  "Upregulated (n=",
  sum(deg_tbl$logFC > log2(FC_threshold) & deg_tbl$adj.P.Val < P_threshold),
  ")"
)

names(keyvals.colour_rd)[keyvals.colour_rd == 'grey30'] <- 'Non-significant'
```

### Build the volcano plot

```{r volcano-enhanced, fig.width=6, fig.height=7}
volcano_plot <- EnhancedVolcano(
  deg_tbl,
  lab = NA,
  x = 'logFC',
  y = 'adj.P.Val',
  title = '',
  subtitle = '',
  pCutoff = P_threshold,
  FCcutoff = log2(FC_threshold),
  gridlines.major = FALSE,
  gridlines.minor = FALSE,
  ylim = c(0, max(-log10(deg_tbl$adj.P.Val), na.rm = TRUE) + 0.5),
  colCustom = keyvals.colour_rd,
  axisLabSize = 20,
  labSize = 3,
  legendLabSize = 8,
  legendIconSize = 7,
  captionLabSize = 16,
  colAlpha = 1,
  pointSize = 0.3
)

volcano_plot

# Save your plot
ggsave(
  "DEG_Volcano_Plot.png",
  volcano_plot,
  device = "png",
  units = "cm",
  width = 32,
  height = 32
)
```

## Method 2: Custom Volcano Plots with ggplot2

When you need finer control — especially to highlight specific genes —
`ggplot2` gives you the flexibility to build it from scratch.

> **Note:** This section assumes `deg_tbl`'s gene identifier column is
> named `V1` (the default when `fwrite()`/`fread()` round-trips a table
> without an explicit row-name column, as in Part 3). If your gene ID
> column has a different name, replace `V1` below with the correct
> column name throughout.

```{r volcano-custom, fig.width=7, fig.height=6}
# Prepare data
deg_tbl$neg_log10_adj.P.Val <- -log10(deg_tbl$adj.P.Val)
deg_tbl$significance <- ifelse(
  deg_tbl$adj.P.Val < P_threshold & abs(deg_tbl$logFC) > log2(FC_threshold),
  ifelse(deg_tbl$logFC > 0, "Upregulated", "Downregulated"),
  "Non-Significant"
)

# 1. Order the entire table by statistical significance first
ordered_tbl <- deg_tbl[order(deg_tbl$adj.P.Val), ]

# 2. Extract the top 15 upregulated genes (logFC > 0)
up_genes <- ordered_tbl$V1[ordered_tbl$logFC > 0][1:15]

# 3. Extract the top 15 downregulated genes (logFC < 0)
down_genes <- ordered_tbl$V1[ordered_tbl$logFC < 0][1:15]

# 4. Combine them into a single vector for labeling
genes_to_label <- c(up_genes, down_genes)

# Create enhanced volcano plot
custom_volcano <- ggplot(deg_tbl, aes(x = logFC, y = neg_log10_adj.P.Val)) +
  geom_point(aes(color = significance), alpha = 0.8, size = 1) +
  scale_color_manual(
    values = c("Upregulated" = "red", "Downregulated" = "blue", "Non-Significant" = "gray80")
  ) +
  geom_vline(xintercept = c(-log2(FC_threshold), log2(FC_threshold)),
             linetype = "dashed", color = "gray50") +
  geom_hline(yintercept = -log10(P_threshold),
             linetype = "dashed", color = "gray50") +
  geom_text_repel(
    data = subset(deg_tbl, V1 %in% genes_to_label),
    aes(label = V1),
    size = 3,
    box.padding = 0.5,
    max.overlaps = Inf
  ) +
  labs(
    x = expression(log[2] ~ "Fold Change"),
    y = expression(-log[10] ~ "adjusted p-value"),
    color = "Differential Expression"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    text = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

custom_volcano
```

## Top Differentially Expressed Genes Table

This table shows the strongest 15 upregulated and 15 downregulated genes
based on adjusted p-value within each direction.

```{r top-deg-table, results='asis'}
gene_id_col <- if ("V1" %in% names(deg_tbl)) "V1" else names(deg_tbl)[1]

top_up_genes <- deg_tbl %>%
  filter(logFC > 0) %>%
  arrange(adj.P.Val) %>%
  slice_head(n = 15) %>%
  transmute(
    Direction = "Upregulated",
    Gene = .data[[gene_id_col]],
    logFC = round(logFC, 3),
    adj.P.Val = signif(adj.P.Val, 3),
    AveExpr = round(AveExpr, 3)
  )

top_down_genes <- deg_tbl %>%
  filter(logFC < 0) %>%
  arrange(adj.P.Val) %>%
  slice_head(n = 15) %>%
  transmute(
    Direction = "Downregulated",
    Gene = .data[[gene_id_col]],
    logFC = round(logFC, 3),
    adj.P.Val = signif(adj.P.Val, 3),
    AveExpr = round(AveExpr, 3)
  )

top_deg_table <- bind_rows(top_up_genes, top_down_genes)

knitr::kable(
  top_deg_table,
  caption = "Top 15 Upregulated and Top 15 Downregulated Genes",
  align = "lrrrr"
)
```

# Creating Informative Heatmaps

Heatmaps are excellent for visualizing expression patterns across
multiple genes and samples, revealing gene clusters that might not be
obvious from other plots.

```{r heatmap, fig.width=6, fig.height=6}
# Prepare sample annotations
col_annot_df <- data.frame(
  Treatment = dge_v$targets$Treatment,
  row.names = rownames(dge_v$targets)
)
col_annot_df <- col_annot_df[order(col_annot_df$Treatment), , drop = FALSE]

# Select significant DEGs
deg_sig <- deg_tbl[abs(deg_tbl$logFC) > log2(FC_threshold) &
                      deg_tbl$adj.P.Val < P_threshold]$V1
expr_deg <- dge_v$E[deg_sig, rownames(col_annot_df), drop = FALSE]

# Create heatmap
heatmap_custom <- pheatmap(
  expr_deg,
  scale = "row",
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  show_colnames = TRUE,
  color = colorRampPalette(c("navy", "white", "red"))(100),
  annotation_col = col_annot_df,
  main = "Differential Expression Heatmap"
)
```

## Alternative Heatmap with ComplexHeatmap

This version gives you more control over clustering, annotations, and
publication styling.

```{r complexheatmap, fig.width=7, fig.height=7}
if (!requireNamespace("ComplexHeatmap", quietly = TRUE) ||
    !requireNamespace("circlize", quietly = TRUE)) {
  message("Install ComplexHeatmap and circlize to render this heatmap block.")
} else {
# Row-scale the expression matrix for visual comparison across samples
expr_deg_scaled <- t(scale(t(expr_deg)))
expr_deg_scaled[is.na(expr_deg_scaled)] <- 0

# Build a named color map for the treatment annotation
treatment_levels <- unique(col_annot_df$Treatment)
treatment_colors <- setNames(c("steelblue", "tomato")[seq_along(treatment_levels)], treatment_levels)

sample_annotation <- ComplexHeatmap::HeatmapAnnotation(
  Treatment = col_annot_df$Treatment,
  col = list(Treatment = treatment_colors)
)

heatmap_colors <- circlize::colorRamp2(
  c(min(expr_deg_scaled, na.rm = TRUE), 0, max(expr_deg_scaled, na.rm = TRUE)),
  c("steelblue", "white", "tomato")
)

heatmap_complex <- ComplexHeatmap::Heatmap(
  expr_deg_scaled,
  name = "Z-score",
  col = heatmap_colors,
  top_annotation = sample_annotation,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  show_column_names = TRUE,
  row_dend_side = "left",
  column_split = col_annot_df$Treatment,
  column_title = "Differential Expression Heatmap",
  heatmap_legend_param = list(title = "Row Z-score")
)

ComplexHeatmap::draw(
  heatmap_complex,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)
}
```

### Saving the heatmap to PDF

`pheatmap` objects need a small helper function to save cleanly to PDF:

```{r save-heatmap, eval=FALSE}
save_pheatmap_pdf <- function(x, filename, width = 8, height = 8) {
  stopifnot(!missing(x))
  stopifnot(!missing(filename))
  pdf(filename, width = width, height = height)
  grid::grid.newpage()
  grid::grid.draw(x$gtable)
  dev.off()
}

save_pheatmap_pdf(heatmap_custom, "heatmap_custom.pdf", width = 4, height = 4)
```

# Visualizing Individual Gene Expression

Sometimes you need to focus on a single gene of interest. Here are three
effective ways to do that.

```{r gene-setup}
gene_to_plot <- "Dclk1"  # Replace with your gene of interest

# Transform all gene expression values to positive (needed for some plot types)
dge_v$E <- dge_v$E + abs(min(dge_v$E)) + 1

gene_data <- data.frame(
  Treatment = dge_v$targets$Treatment,
  Expression = dge_v$E[gene_to_plot, ]
)

# Keep treatment order stable in the plots
gene_data$Treatment <- factor(gene_data$Treatment, levels = unique(gene_data$Treatment))

# Quick sanity check: inspect the mean expression by group before interpreting direction
gene_summary <- gene_data %>%
  group_by(Treatment) %>%
  summarise(mean_expression = mean(Expression), .groups = "drop")

gene_summary
```

## 1. Violin Plots: Showing Distribution and Individual Points

```{r violin-plot, fig.width=5, fig.height=5}
violin_plot <- ggplot(gene_data, aes(x = Treatment, y = Expression)) +
  geom_violin(aes(fill = Treatment), alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 1.0) +
  stat_summary(fun = mean, geom = "point", size = 3, color = "black") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  scale_fill_manual(values = c(treatment_colors[2], treatment_colors[1])) +
  theme_minimal() + 
  labs(title = paste(gene_to_plot, "Expression"), y = "log2 Expression") +
  theme(legend.position = "none")

violin_plot
```

## 2. Box Plots: Traditional and Clear

```{r box-plot, fig.width=5, fig.height=5}
boxplot <- ggplot(gene_data, aes(x = Treatment, y = Expression)) + 
  geom_boxplot(aes(fill = Treatment), alpha = 0.7) + 
  geom_jitter(width = 0.2, alpha = 0.6) + 
  scale_fill_manual(values = c(treatment_colors[2], treatment_colors[1])) +
  theme_minimal() + 
  labs(title = paste(gene_to_plot, "Expression"), y = "log2 Expression") + 
  theme(legend.position = "none") 

boxplot
```

## 3. Bar Plots: Showing Means with Error Bars

```{r bar-plot, fig.width=5, fig.height=5}
# Calculate means and standard errors
summary_data <- gene_data %>%
  group_by(Treatment) %>%
  summarise(
    mean = mean(Expression),
    se = sd(Expression) / sqrt(n())
  )

# Create bar plot
bar_plot <- ggplot(summary_data, aes(x = Treatment, y = mean, fill = Treatment)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  scale_fill_manual(values = c(treatment_colors[2], treatment_colors[1])) +
  theme_minimal() + 
  labs(title = paste(gene_to_plot, "Expression"), y = "Mean log2 Expression") +
  theme(legend.position = "none")

bar_plot
```

# Best Practices for RNA-seq Visualization

**Choose the right plot type**

- Volcano plots — global differential expression overview
- Heatmaps — expression patterns across multiple genes
- Violin/box plots — detailed views of individual genes

**Color selection**

- Use colorblind-friendly palettes
- Maintain consistency across related figures
- Use appropriate color scales for different data types

**Labels and annotations**

- Include clear axis labels
- Add informative titles
- Label key features or genes of interest

**Statistical information**

- Show significance thresholds
- Include error bars where appropriate
- Indicate sample sizes

# Conclusion

Effective visualization is essential for understanding and communicating
RNA-seq results. The techniques above — volcano plots, heatmaps, and
individual gene expression plots — form a solid foundation for building
clear, informative, publication-ready figures.

# Next Steps

The next tutorial in this series covers pathway analysis, exploring the
biological mechanisms underlying your differential expression results.
