---
title: "RNAseq Analysis: From DEGs to Pathways"
subtitle: "Beginner's Workflow (Part 5) — GO, KEGG, and MSigDB Enrichment (GSEA and ORA)"
author: "Dr. Lei Guo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

After data preparation, statistical testing, and visualization (Parts
3–4), the next step is to explore the biological significance of the
differentially expressed genes (DEGs) — this is functional enrichment /
pathway analysis.

Pathway analysis helps you:

- Connect expression changes to known biological mechanisms
- Provide biological context for experimental findings
- Generate hypotheses for further investigation

# Pathway Databases

## Gene Ontology (GO)

A structured vocabulary describing gene functions, processes, and
cellular locations, organized into three categories:

- **Molecular Function (MF)** — specific molecular-level activities
  (e.g. ATP binding, protein kinase activity)
- **Biological Process (BP)** — broader biological goals involving
  multiple molecular functions (e.g. cell cycle regulation, immune
  response, apoptosis)
- **Cellular Component (CC)** — where gene products are active within
  the cell (e.g. nucleus, cytoplasm, mitochondrial membrane)

## KEGG (Kyoto Encyclopedia of Genes and Genomes)

Integrates genomic, chemical, and functional information:

- **Pathways** — detailed maps of biological processes
- **KEGG Orthology (KO)** — groups of genes with shared function across
  species
- **KEGG Disease** — disease pathways and gene–disease associations
- **KEGG Drug and Enzyme** — drug interactions and metabolic enzyme
  information
- **KEGG Compound** — chemical compound structures and roles

## Molecular Signatures Database (MSigDB)

The most comprehensive pathway database, incorporating GO, KEGG, and
more:

- **Hallmark Gene Sets** — core biological processes and pathways
- **Curated Gene Sets (C2)** — pathways from KEGG, Reactome, and
  published studies
- **Motif Gene Sets (C3)** — gene regulatory motifs
- **Computational Gene Sets (C4)** — co-expression networks
- **GO Gene Sets (C5)** — Gene Ontology groupings
- **Oncogenic Signatures (C6)** — cancer-related pathways
- **Immunologic Signatures (C7)** — immune system patterns
- **Cell Type Signatures (C8)** — cell-specific expression patterns

# Statistical Methods for Pathway Analysis

## 1. Over-Representation Analysis (ORA)

Uses statistical tests (typically Fisher's exact test or hypergeometric
test) to evaluate gene set enrichment.

**Advantages:** simple to understand and implement.

**Limitations:** requires predefined cutoffs for DEG selection, treats
all genes as equally important, doesn't consider expression change
magnitude.

## 2. Gene Set Enrichment Analysis (GSEA)

Uses a running-sum statistic to analyze gene distribution patterns
within a ranked list.

**Advantages:** no arbitrary cutoffs needed, considers expression change
magnitude, detects subtle expression patterns.

**Limitations:** more computationally intensive, can be more complex to
interpret.

# Practical Implementation

## Prerequisites

```{r install-packages, eval=FALSE}
# Install required packages if needed
install.packages(c("BiocManager"))
BiocManager::install(c("clusterProfiler", "org.Hs.eg.db", "org.Mm.eg.db",
                        "org.Rn.eg.db", "pathview", "enrichplot", "DOSE"))

install.packages(c("data.table", "ggplot2"))
```

## 1. Data Preparation

### 1.1 Loading Required Libraries

```{r load-packages}
library(data.table)
library(clusterProfiler)
# library(org.Hs.eg.db)
library(org.Mm.eg.db)
# library(org.Rn.eg.db)
library(ggplot2)
library(DOSE)
library(enrichplot)
```

Uncomment `org.Hs.eg.db` (human) or `org.Rn.eg.db` (rat) instead of
`org.Mm.eg.db` (mouse) depending on your organism.

### 1.2 Input Data Format

Import the DEG table created in Part 3. For ORA, you need P-value and
fold-change thresholds to filter DEGs. **GSEA does not require
pre-filtering** — it analyzes the entire ranked gene list.

```{r load-deg}
deg <- fread("DEG_P1_lfc0.tsv")

p_threshold <- 0.05
fc_threshold <- 2
```

> **Note:** Throughout this tutorial, `deg$V1` refers to the gene
> identifier column. If your gene ID column has a different name (e.g.
> `Gene`), substitute it wherever `V1` appears below.

## 2. Gene Set Enrichment Analysis (GSEA)

### 2.1 GO Enrichment Analysis

```{r gsea-go}
# rank the DEGs by fold change
deg_order_fc <- deg[order(-logFC)] # rank the genes by logFC, descending
logfc <- deg_order_fc$logFC        # get logFC
names(logfc) <- deg_order_fc$V1    # make a named vector of logFC

# GSEA
# Mouse: OrgDb = org.Mm.eg.db; Rat: OrgDb = org.Rn.eg.db; Human: OrgDb = org.Hs.eg.db
enrich_go_gsea <- gseGO(geneList = logfc, OrgDb = org.Mm.eg.db, ont = "ALL",
                         pvalueCutoff = 0.05, keyType = "SYMBOL", verbose = FALSE)

# save the enrichment table
enrich_go_gsea_df <- enrich_go_gsea@result

fwrite(enrich_go_gsea_df, "enrich_go_gsea_df.tsv", sep = "\t")
```

### 2.2 Visualize the Top Enriched GO Terms

The `clusterProfiler` package offers many visualization options. We
start with a dot plot — the most widely used and intuitive
representation. See the `clusterProfiler` documentation for more
advanced options.

```{r gsea-go-dotplot, fig.width=8, fig.height=8}
dotplot_enrich_go_gsea <- dotplot(enrich_go_gsea, showCategory = 10, orderBy = "GeneRatio")

dotplot_enrich_go_gsea

ggsave("dotplot_enrich_go_gsea.png", dotplot_enrich_go_gsea,
       device = "png", units = "cm", width = 16, height = 18)
```

GSEA results can also be visualized at the individual pathway level:

```{r gsea-go-single-plot, fig.width=8, fig.height=7}
# specify the GO term you want to plot (check enrich_go_gsea_df above)
GOID <- "GO:0010634"

gsea_plot <- gseaplot2(enrich_go_gsea, geneSetID = GOID,
                        title = subset(enrich_go_gsea_df, ID == GOID)$Description)

gsea_plot

ggsave(paste0("gsea_plot_", gsub(":", "", GOID), ".png"), gsea_plot,
       device = "png", units = "cm", width = 18, height = 16)
```

> Replace `GOID` with any GO term ID you're interested in from your own
> `enrich_go_gsea_df` results.

### 2.3 KEGG Enrichment Analysis

Same workflow as above, but uses `gseKEGG()`. **KEGG requires Entrez
IDs, not gene symbols**, so gene symbols must be converted first. Note
the `organism` codes: `"hsa"` (human), `"mmu"` (mouse), `"rno"` (rat).

```{r kegg-setup}
# Load required annotation packages
# Choose the organism-specific database that matches your dataset:
#   - org.Mm.eg.db : Mouse
#   - org.Hs.eg.db : Human
#   - org.Rn.eg.db : Rat
library(AnnotationDbi)
library(org.Mm.eg.db)
library(org.Rn.eg.db)
library(org.Hs.eg.db)
```

```{r kegg-convert-ids}
# ------------------------------------------------------------
# 1. Convert gene symbols -> Entrez IDs using mapIds()
# ------------------------------------------------------------
# 'logfc' should be a named numeric vector where names = gene symbols.
# We convert those gene symbols to Entrez IDs because KEGG requires Entrez IDs.
entrez_ids <- mapIds(
  x = org.Mm.eg.db,        # Change to org.Hs.eg.db (human) or org.Rn.eg.db (rat) as needed
  keys = names(logfc),     # Input gene symbols
  keytype = "SYMBOL",      # Type of input identifiers
  column = "ENTREZID",     # Type of output identifiers
  multiVals = "first"      # If multiple Entrez IDs map to one symbol, keep the first
)

# Replace gene symbol names with Entrez IDs
names(logfc) <- entrez_ids

# Remove genes that failed to map (NA Entrez IDs)
logfc <- logfc[!is.na(names(logfc))]
```

```{r kegg-gsea, eval=FALSE}
# ------------------------------------------------------------
# 2. Run KEGG GSEA (gseKEGG) for different organisms
# ------------------------------------------------------------
# KEGG organism codes:
#   - "mmu" : Mouse
#   - "hsa" : Human
#   - "rno" : Rat

# Mouse KEGG GSEA
enrich_kegg_gsea <- gseKEGG(
  geneList = logfc,
  organism = "mmu"
)

# Rat KEGG GSEA
enrich_kegg_gsea <- gseKEGG(
  geneList = logfc,
  organism = "rno"
)

# Human KEGG GSEA
enrich_kegg_gsea <- gseKEGG(
  geneList = logfc,
  organism = "hsa"
)
```

> Run only the block matching your organism — the three calls above are
> alternatives, not sequential steps (each one overwrites
> `enrich_kegg_gsea`).

## 3. Over-Representation Analysis (ORA)

### 3.1 GO Enrichment Analysis

ORA requires only a list of DEGs as input. This example focuses on
up-regulated genes; the same process applies to down-regulated genes.

```{r ora-go}
# make a gene list
deg_up <- deg[logFC > log2(fc_threshold) & adj.P.Val < p_threshold]$V1
# deg_dn <- deg[logFC < -log2(fc_threshold) & adj.P.Val < p_threshold]$V1

enrich_go_fet_up <- enrichGO(gene = deg_up, OrgDb = org.Mm.eg.db, keyType = "SYMBOL",
                              ont = "ALL", pvalueCutoff = 0.05, pAdjustMethod = "BH",
                              qvalueCutoff = 0.05)

enrich_go_fet_up_df <- enrich_go_fet_up@result

fwrite(enrich_go_fet_up_df, "enrich_go_fet_up_df.tsv", sep = "\t")
```

### 3.2 Visualize the Top Enriched GO Terms

```{r ora-go-dotplot, fig.width=8, fig.height=8}
# show the top 10 terms
dotplot_enrich_go_fet_up <- dotplot(enrich_go_fet_up, showCategory = 10, orderBy = "GeneRatio")

dotplot_enrich_go_fet_up

ggsave("dotplot_enrich_go_fet_up.png", dotplot_enrich_go_fet_up,
       device = "png", units = "cm", width = 16, height = 18)
```

### 3.3 KEGG Enrichment Analysis

Same methodology as the GO enrichment above, but uses `enrichKEGG()`.
Convert gene symbols to Entrez IDs first (see the [KEGG ID conversion
step](#kegg-setup) above) — `enrichKEGG()` requires Entrez IDs, not
symbols.

```{r ora-kegg, eval=FALSE}
enrich_kegg_fet_up <- enrichKEGG(
  gene     = deg_up,     # Vector of Entrez IDs
  organism = "mmu",      # Change to "hsa" (human) or "rno" (rat) as needed
  keyType  = "kegg"      # Required when using Entrez IDs
)
```

> **Important:** `deg_up` above is a vector of gene *symbols* as defined
> in section 3.1. Before running `enrichKEGG()`, convert it to Entrez
> IDs the same way `logfc` was converted in section 2.3, e.g.:
>
> ``` r
> deg_up_entrez <- mapIds(org.Mm.eg.db, keys = deg_up, keytype = "SYMBOL",
>                          column = "ENTREZID", multiVals = "first")
> deg_up_entrez <- na.omit(deg_up_entrez)
> ```
>
> then pass `gene = deg_up_entrez` to `enrichKEGG()`.

## 4. MSigDB Enrichment

The most comprehensive option — incorporates GO, KEGG, Hallmark,
Reactome, and other pathway collections in one analysis.

Before starting, download the appropriate gene set `.gmt` file from the
MSigDB website (choose the species-specific file — mouse or human), and
place it in your working directory or note its full path.

### 4.1 The GSEA Method

```{r msigdb-gsea, eval=FALSE}
msigdb <- read.gmt("msigdb.v2025.1.Mm.symbols.gmt")

enrich_msigdb_gsea <- GSEA(logfc, TERM2GENE = msigdb)

enrich_msigdb_gsea_df <- enrich_msigdb_gsea@result

fwrite(enrich_msigdb_gsea_df, "enrich_msigdb_gsea_df.tsv", sep = "\t")
```

### 4.2 The ORA Method

```{r msigdb-ora, eval=FALSE}
msigdb <- read.gmt("msigdb.v2025.1.Mm.symbols.gmt")

enrich_msigdb_fet_up <- enricher(deg_up, TERM2GENE = msigdb)

enrich_msigdb_fet_up_df <- enrich_msigdb_fet_up@result

fwrite(enrich_msigdb_fet_up_df, "enrich_msigdb_fet_up_df.tsv", sep = "\t")
```

> Update the `.gmt` filename to match the species and MSigDB version you
> downloaded (e.g. `msigdb.v2025.1.Hs.symbols.gmt` for human).

# Results Interpretation

**Statistical significance**

- Examine adjusted p-values (FDR)
- Consider gene set sizes
- Look for consistent patterns across methods

**Biological relevance**

- Focus on pathways related to your experimental context
- Consider pathway relationships and hierarchies
- Look for unexpected but interesting connections

**Technical considerations**

- Account for gene set size effects
- Verify multiple testing correction methods
- Ensure appropriate background gene selection

# Best Practices

- Use appropriate statistical cutoffs consistently
- Consider your experimental context when interpreting results
- Validate key findings using alternative methods
- Document all analysis parameters and software versions
- Cross-reference findings with current literature
- Consider pathway redundancy and overlap

# Troubleshooting Notes

- **`gseKEGG()`/`enrichKEGG()` error about needing Entrez IDs** — GO
  functions (`gseGO`, `enrichGO`) accept gene symbols directly via
  `keyType = "SYMBOL"`, but KEGG functions always require Entrez IDs.
  Convert symbols to Entrez IDs with `mapIds()` (see section 2.3) before
  calling any KEGG function.
- **`NA` values after `mapIds()`** — not every gene symbol has a
  matching Entrez ID. Always filter these out with
  `logfc <- logfc[!is.na(names(logfc))]` (or the equivalent for a plain
  vector) before running KEGG functions.
- **Wrong organism database** — double-check that `OrgDb` (GO) and
  `organism` (KEGG) match your actual species; mismatches will silently
  return few or no enriched terms rather than an error.

# References

- Yu G, Wang LG, Han Y, He QY (2012). clusterProfiler: an R package for
  comparing biological themes among gene clusters. *OMICS: A Journal of
  Integrative Biology.*
- Love MI, Huber W, Anders S (2014). Moderated estimation of fold change
  and dispersion for RNA-seq data with DESeq2. *Genome Biology.*

# Additional Resources

- ClusterProfiler Documentation
- GSEA Homepage
- Bioconductor Workflow
